---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
    echo: true
    message: false
    warning: false
    tidy: true
---

Some notes on this Rmd file are available in the file NOTES.md

## Loading required libraries and packages
```{r, include = TRUE}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
})
```


## Loading and preprocessing the data
```{r}
if (!file.exists("activity.csv")) {
  if (!file.exists("activity.zip")) {
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", "activity.zip", method = "libcurl")
  }
  unzip("activity.zip")
}
activity <- read.csv("activity.csv", header = TRUE, stringsAsFactors = FALSE)
activity$date <- as.Date(activity$date)
head(activity)
```


## What is mean total number of steps taken per day?

First I calculate the number of steps taken each day:

```{r}
steps_per_day <- activity %>% group_by(date) %>% summarise(steps = sum(steps, na.rm = T))
mean_steps_day <- mean(steps_per_day$steps)
median_steps_day <- median(steps_per_day$steps)
```

The figure below shows the histogram of the steps taken each day, the mean value (`r mean_steps_day` steps, red) and the median value (`r median_steps_day` steps, blue)

```{r steps_per_day_histogram}
p <- ggplot(steps_per_day, aes(x=steps)) + geom_histogram(binwidth = 1500) +
  geom_vline(xintercept = mean_steps_day, color="red") +
  geom_vline(xintercept = median_steps_day, color="blue") 
p
```

## What is the average daily activity pattern?

```{r}
interval_activity <- activity %>% group_by(interval) %>% summarise(steps = mean(steps, na.rm=T))
interval_max_steps <- interval_activity$interval[which.max(interval_activity$steps)]
```

The interval with the average number of steps across all days is `r interval_max_steps`.

```{r avg_steps_interval}
p <- ggplot(interval_activity, aes(x=interval, y = steps)) + geom_line()
p
```

## Imputing missing values

```{r}
n_rows_with_NA <- sum(ifelse(is.na(activity$steps),1,0))
```
The original dataset contains `r n_rows_with_NA` rows with missing values.

I imputed the value by using the average value in the same interval (which I already had to compute for the previous step):

```{r}
activity_imputed <- activity %>% left_join(select(interval_activity, interval, avg = steps), by = "interval") %>%
  mutate(steps_imputed = ifelse(is.na(steps), avg, steps))
head(activity_imputed)
```

```{r}
steps_per_day_imputed <- activity_imputed %>% group_by(date) %>% summarise(steps = sum(steps_imputed))
mean_steps_day_imputed <- mean(steps_per_day_imputed$steps)
median_steps_day_imputed <- median(steps_per_day_imputed$steps)
```

The figure below show the histogram of steps per day when imputing the number of steps using the average value in the interval,
the mean value (`r round(mean_steps_day_imputed,2)` steps, red line) and the median_value (`r round(median_steps_day_imputed,2)` steps, blue line):

```{r steps_per_day_imputed_histogram}
p <- ggplot(steps_per_day_imputed, aes(x=steps)) + geom_histogram(binwidth = 1500) +
  geom_vline(xintercept = mean_steps_day_imputed, color="red") +
  geom_vline(xintercept = median_steps_day, color="blue") 
p
```

## Are there differences in activity patterns between weekdays and weekends?

The time series show that during weekend the rise in steps starts usually at larger intervals
(presumably because people wake up later?), and the peak in steps before interval 1000 is lower.

However after the interval 1000 the number of steps per interval is usually larger during weekends
than weekdays (presumably because people take a walk, go shopping, do sports etc).

```{r}
ts_activity_imputed <- activity_imputed %>% mutate(day_type = factor(weekdays(date) %in% c("Saturday","Sunday"), 
                                              levels = c(TRUE, FALSE), labels = c("weekend","weekday"))) %>%
  group_by(day_type, interval) %>% summarise(steps = mean(steps_imputed)) %>% ungroup() %>%
  arrange(day_type, interval)

head(ts_activity_imputed)
```

```{r average_activity_date_day_type}
p <- ggplot(ts_activity_imputed, aes(x = interval, y = steps)) + geom_line() + 
  facet_grid(day_type ~ .)
p
```
